---
title: "Meta-analysis"
page-layout: full
title-block-banner: true
format: 
  html: 
    theme: lux 
    css: styles/meta_analysis.css
    toc: true
    toc-depth: 3
    toc-title: "**Table of contents**"
editor: 
  markdown: 
    wrap: 72
---

Meta-analysis is a statistical approach for combining results from multiple independent studies to estimate an overall effect and identify factors that explain variation among studies.

## What is a meta-analysis?

Ecological studies often report different results because they are conducted in different environments, use different methods, or examine different species and systems. Meta-analysis provides a quantitative framework for synthesizing these findings.

A typical meta-analysis involves:

1. **Defining a research question**
2. **Systematically searching the literature**
3. **Screening and selecting relevant studies**
4. **Extracting data from individual studies**
5. **Calculating effect sizes**
6. **Fitting meta-analytic models**
7. **Examining heterogeneity and moderators**
8. **Assessing the robustness of the results**

## Effect sizes

An effect size provides a common metric that allows results from different studies to be compared. For continuous ecological responses, the log response ratio (lnRR) is commonly used:

\[
\mathrm{lnRR} = \ln\left(\frac{\bar{X}_{T}}{\bar{X}_{C}}\right)
\]

where \(\bar{X}_{T}\) is the mean response in the treatment group and \(\bar{X}_{C}\) is the mean response in the control group.

Positive values indicate higher values in the treatment relative to the control, whereas negative values indicate lower values.

## Meta-analytic models

A basic meta-analysis estimates an overall effect across studies. However, ecological datasets commonly contain multiple observations from the same study, site, or experiment. Multilevel meta-analysis can account for this non-independence by including appropriate random effects.

For example:

```{r  eval=FALSE}
library(metafor)

model <- rma.mv(
  yi,
  vi,
  random = ~ 1 | StudyID/CohortID,
  method = "REML",
  data = dat
)

summary(model)
```

## Exploring moderators

The effectiveness of an ecological treatment may depend on environmental conditions, species characteristics, or management practices. These factors can be examined using moderator analyses.

```{r eval=FALSE}
model_mod <- rma.mv(
  yi,
  vi,
  mods = ~ Habitat + MAT + MAP,
  random = ~ 1 | StudyID/CohortID,
  method = "REML",
  data = dat
)

summary(model_mod)
```

Moderator analyses can help identify **when, where, and under what conditions an ecological intervention is most effective**.

## Visualizing results

Forest plots provide a useful way to visualize individual study effects and the overall meta-analytic estimate.

```{r  eval=FALSE}
forest(model)
```

A well-designed meta-analysis should combine rigorous study selection, transparent data extraction, appropriate statistical models, and clear visualization of the results.